Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-03-01 src/tex/ms.tex
Some comments from Will.

\newcommand*{\mi}[1]{\textsf{\color{magenta} [\textbf{MAX:} #1]}}

@@

We carry out \ac{PE} using a custom version of \textsc{Bilby} \citep{Bilby}, that applies Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline waveform.
We take the \ac{PE} configuration as in \citet{GWTC-2.1, GWTC-3, GWTC-2.1_dataset, GWTC-3_dataset} as a starting point, with \textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb} as the reference waveform.
We apply a distance prior which is uniform in comoving volume and source frame time instead of a power law distribution used in \citet{GWTC-2.1, GWTC-3, GWTC-2.1_dataset, GWTC-3_dataset}, and set the prior on $\kappa$ to a uniform distribution between $-1$ and $1$.

@@

Since birefringence is a property of spacetime and we are targeting a propagation effect not intrinsic to any source, we should expect $\kappa$ to take the same value for all signals, whether it vanishes or not.
\mi{caveats?}
Under this assumption, the posterior on $\kappa$ inferred collectively from all events with a uniform prior is simply obtained from the product of the individual likelihoods, $p(d_i \mid \kappa) \propto p(\kappa \mid d_i)/p(\kappa)$, such that
\begin{equation}
    p(\kappa \mid \{d_i\})\propto p(\kappa) \prod_{i}\frac{p(\kappa \mid d_i)}{p(\kappa)}\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event, and $p(\kappa)$ is the prior on $\kappa$; since the prior is uniform, in our case Eq.~\eqref{eq:restricted_posterior} reduces to the product of the posteriors, namely $p(\kappa \mid \{d_i\}) \propto \prod_{i}p(\kappa \mid d_i)$.
We use Eq.~\eqref{eq:restricted_posterior} to obtain the primary constraint presented in this work.

@@

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal further physics than is implied by Eq.~\eqref{eq:waveform_modification} \mi{which physics? @Will, any ideas?} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.

>>>>

\newcommand*{\mi}[1]{\textsf{\color{magenta} [\textbf{MAX:} #1]}}
\newcommand*{\wf}[1]{\textsf{\color{cyan} [\textbf{WILL:} #1]}}

@@

We carry out \ac{PE} using a custom version of \textsc{Bilby} \citep{Bilby},
that applies Eq.~\eqref{eq:waveform_modification} for any \ac{GR} baseline
waveform. We take the \ac{PE} configuration as in \citet{GWTC-2.1, GWTC-3,
GWTC-2.1_dataset, GWTC-3_dataset} as a starting point, with
\textsc{IMRPhenomXPHM} \citep{Pratten:2020ceb} as the reference waveform. We
apply a distance prior which is uniform in comoving volume and source frame time
instead of a power law distribution used in \citet{GWTC-2.1, GWTC-3,
GWTC-2.1_dataset, GWTC-3_dataset}, and set the prior on $\kappa$ to a uniform
distribution between $-1$ and $1$.  \wf{Actually, the LIGO references
\emph{also} used the ``cosmological'' prior that is uniform merger rate density
in the source frame; maybe just drop the ``instead of\ldots''?}

@@

Since birefringence is a property of spacetime and we are targeting a
propagation effect not intrinsic to any source, we should expect $\kappa$ to
take the same value for all signals, whether it vanishes or not. \mi{caveats?}
\wf{\emph{If} birefringence were a property of spacetime, this would be true;
but in (some? most? all?) ``parity violating'' theories, the parity violation is
due to an extra \emph{field} that is parity-odd and couples to gravity.  In this
case whether the birefringence is the same along the different lines of sight to
the different events depends on \emph{whatever that field is doing between us
and the events}.  The \emph{simplest} assumption is that the field is ``the
same'' or shared from event-to-event, but it's by no means required that it be
so.  Hence the analysis of the next section.} Under this assumption, the
posterior on $\kappa$ inferred collectively from all events with a uniform prior
is simply obtained from the product of the individual likelihoods, $p(d_i \mid
\kappa) \propto p(\kappa \mid d_i)/p(\kappa)$, such that
\begin{equation}
    p(\kappa \mid \{d_i\})\propto p(\kappa) \prod_{i}\frac{p(\kappa \mid d_i)}{p(\kappa)}\,,
    \label{eq:restricted_posterior}
\end{equation}
where $d_i$ is the strain data for the $i$\textsuperscript{th} event, and $p(\kappa)$ is the prior on $\kappa$; since the prior is uniform, in our case Eq.~\eqref{eq:restricted_posterior} reduces to the product of the posteriors, namely $p(\kappa \mid \{d_i\}) \propto \prod_{i}p(\kappa \mid d_i)$.
We use Eq.~\eqref{eq:restricted_posterior} to obtain the primary constraint presented in this work.

@@

If \ac{GR} is correct and there is no birefringence, we should find the observed $\kappa$ distribution to be consistent with a delta function at the origin ($\kappa_i = 0$ for all $i$, or $\mu=\sigma=0$); on the other hand, if spacetime is birefringent, we expect to find a delta function at some nonzero value ($\kappa_i = \kappa \neq 0$, or $\mu = \kappa$ and $\sigma=0$).
But this analysis also has the power to reveal unexpected physics or systematics in our measurements: if $\sigma$ is confidently found to be nonzero, this would imply that our set of measurements is statistically unlikely to originate from a unique $\kappa$ value.
This could signal further physics than is implied by Eq.~\eqref{eq:waveform_modification} \mi{which physics? @Will, any ideas?} \wf{see comment above} or, more prosaically, that there are outliers in our measurements due to mismodeling, e.g., in the waveform approximant or the noise of the detector.
